---
title: "HDS5101"
author: "Phoebe Wong Hui Lei"
date: "2026-03-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown


**Goal of the task** : 
Find the best model in discriminating individuals with or without heart disease by evaluating the brier score, area under the ROC curve and calibration curve. 

**Source of dataset**: https://archive.ics.uci.edu/ml/machine-learning-databases/heart-disease/processed.cleveland.data

**Study population**: Records collected from Cleveland Clinic Foundation

**Procedure**: 
1) Assess the missing variables in the dataset, and the extent of missingness
2) Apply single imputation technique to handle missing data, and create complete dataset for analysis
3) Evaluate differences between complete case dataset and the imputed dataset to assess the impact of imputation 
4) Randomly split the dataset into 70% of train and 30% of test for model development 
5) Fit multiple predictive models on the train dataset and perform prediction on the test dataset
6) Make comparison on models performance using metrics such as Brier score and AUC. Visualize the ROC curve for comparison on the classification accuracy

**Outcome of the study** :  
Presence of heart disease in the patient. The original variable ranges from 0 to 4, representing different levels of heart disease severity. For this study, the outcome variable will be converted into a binary variable where 0 indicates absence of heart disease and values 1–4 indicate presence of heart disease


# Load library
```{r}
library(dplyr)
library(mice)
library(VIM)
library(pROC)
library(CalibrationCurves)
library(glmnet)
library(MASS)
library(e1071)
library(GGally)
library(pls)
library(tree)
library(randomForest)
```


```{r}
url <- "https://archive.ics.uci.edu/ml/machine-learning-databases/heart-disease/processed.cleveland.data"

cleveland <- read.table(url, sep = ",", header = FALSE)

colnames(cleveland)<-c("age","sex","cp","trestbps","chol","fbs","restecg","thalach","exang","oldpeak","slope","ca","thal","num")

# convert variable to factor
cleveland <- cleveland %>%
  mutate(
    sex=factor(sex),
    cp=factor(cp),
    fbs=factor(fbs),
    restecg=factor(restecg),
    exang=factor(exang),
    slope=factor(slope)
  )


head(cleveland)
```

```{r}
library(naniar)

cleveland <- cleveland %>%
  replace_with_na_all(condition = ~.x =="?")

#check for null value
na_count <- colSums(is.na(cleveland))
na_count
```

# Convert multi-category to binary form
```{r}
cleveland <- cleveland %>% 
  mutate(
    heart_disease = factor(ifelse(num >= 1, 1, 0),levels = c(0,1))
  )

cleveland %>%
  count(heart_disease)
```

# Summary of data
```{r}
library(table1)
# summary of dataset
table1(~ age+sex+cp+trestbps+chol+fbs+restecg+thalach+exang+oldpeak+slope+ca+thal| heart_disease ,data=cleveland)
```

# Visualization of Missing Data
```{r}
md.pattern(cleveland, rotate.names=TRUE)
```

```{r}
aggr(cleveland)
```

# Imputation
```{r}
cleveland.to.impute<- cleveland %>%
  dplyr::select(-num)%>%
  mutate(
    thal = factor(thal),
    ca = factor(ca)
  )

# use polyreg since there are more than 2 levels of factor
cleveland.impute <- mice(data=cleveland.to.impute,m=10,method=c("","","","","","","","","","","","polyreg","polyreg",""), seed=100, print=FALSE)
```

```{r}
plot(cleveland.impute)
```


```{r}
#pick the 5th imputed dataset
cleveland_imputed <- complete(cleveland.impute,action=5)
```

# Data Visualization on the outcome
```{r}
ggplot(cleveland_imputed, aes(x = heart_disease,fill=heart_disease)) +
  geom_bar()+labs(
    x="Heart disease",
    y="Count",
    title="Distribution of heart disease"
  )
```

The barplot above shows a distribution of outcome variable determining whether an individual has heart disease or not. It shows that number of observations for individuals without heart disease slightly higher than those with.

# Check for correlation among the continuous predictor
```{r}
con_cleveland_imputed <- cleveland_imputed %>%
  dplyr::select(age,oldpeak,trestbps,chol,thalach)

ggpairs(con_cleveland_imputed,label = TRUE)
```
Based on the correlation heatmap shows that most numeric variables have low. It suggest that there is less risk of multicollinearity.

# Split the dataset into train and test
```{r}
# set the dataset to be split on 70% as train
set.seed(100)

hd <- cleveland_imputed$heart_disease
n <- nrow(cleveland_imputed)
train <- runif(n)< 0.7
train.hd <- cleveland_imputed[train,]
test.hd <- cleveland_imputed[!train,]
dim(test.hd)
hd.test <- hd[!train]
```
## Brier score function
```{r}
brier_score <- function(y, pred_prob) {
  mean((y - pred_prob)^2)
}

```


# GLM - Complete Case Analysis
```{r}
cleveland <- cleveland %>% dplyr::select(-num)

cleveland_cc <- na.omit(cleveland)

n_cc <- nrow(cleveland_cc)
train_idx <- runif(n_cc)< 0.7

train_cc <- cleveland_cc[train_idx,]
test_cc <- cleveland_cc[!train_idx,]

hd_cc.test <- cleveland_cc$heart_disease[!train_idx]

glm.fits.cc <- glm(heart_disease ~ .,data = train_cc, family = binomial)

glm.probs.cc <- predict(glm.fits.cc,newdata=test_cc,type = "response") 
```

## Brier score
```{r}
brier_test_glm_cc <- brier_score(as.numeric(hd_cc.test) - 1, glm.probs.cc)
brier_test_glm_cc
```

The brier score of fitted generalized linear model on complete case is 0.0850.

## ROC curve
```{r}
roc_obj_glm_cc <- roc(as.numeric(hd_cc.test) - 1, glm.probs.cc)
plot(roc_obj_glm_cc, print.auc=TRUE)
```
In the ROC curve, it has shown the trade off between correctly identifying patients with heart disease(TP) and incorrectly identifying patients having heart disease as they do not(FP). 

The AUC shows an overall measure of model discrimination between individuals with and without heart disease. An AUC with value of 0.949 indicates well discrimination.When randomly select an individual with heart disease and one without heart disease,the model assigns higher predicted probability of identifying heart disease to the affected individual 94.9% of the time.

## Calibration curve
```{r}
valProbggplot(glm.probs.cc, as.numeric(hd_cc.test)-1)
```

The calibration intercept of 0.27, which deviates from ideal value of 0, indicating the model is underestimating predicted probability. The slope of 1.22, which is greater than 1,less than 1,suggesting the predicted probabilities are not extreme enough.

# GLM fit - Imputed dataset

```{r}
glm.fits <- glm(heart_disease ~ .,data = train.hd, family = binomial)

glm.probs <- predict(glm.fits,newdata=test.hd,type = "response") ## predict probabilities in the test set
```

## Brier score
```{r}
brier_test_glm <- brier_score(as.numeric(hd.test) - 1, glm.probs)
brier_test_glm
```
The brier score of fitted generalized linear model is 0.1255.

## ROC curve
```{r}
roc_obj_glm <- roc(as.numeric(hd.test) - 1, glm.probs)
plot(roc_obj_glm, print.auc=TRUE)
```
An AUC with value of 0.901 indicates the model assigns higher predicted probability of identifying heart disease to the affected individual 90.1% of the time.


## Calibration curve
```{r}
cali.curve.glm <- valProbggplot(glm.probs, as.numeric(hd.test)-1)
cali.curve.glm
```

The calibration intercept of 0.28, which deviates from ideal value of 0, indicating the model is underestimating predicted probability. The slope of 0.64, which is less than 1,suggesting that the predicted probabilities are too extreme.


**Comparison of models performance on complete case vs imputed case **
```{r}
comparison_cc <- data.frame(
  brier = c(
    glm=brier_test_glm,
    glm_cc=brier_test_glm_cc
    
  ),
  auc = c(
    glm=roc_obj_glm$auc,
    glm_cc=roc_obj_glm_cc$auc
  )
)
round(comparison_cc,4)
```
Based on the comparison of the Brier score and AUC score, there is only a minimal difference between the generalized linear model (GLM) fitted on the imputed dataset and the complete case dataset. The Brier score differs by approximately 0.04, while the AUC score shows a small difference of 0.04 between the two approaches. Given that the model performances are very similar, imputed dataset is used in further analysis to avoid the loss of information and variability that may occur when using complete case analysis due to the missing data.


# Ridge regression
```{r}
set.seed(100)

x <- model.matrix(heart_disease ~ ., cleveland_imputed)[, -1]
y <- cleveland_imputed$heart_disease

ridge.mod<-glmnet(x[train,], y[train], alpha = 0, family = binomial)

cv.out <- cv.glmnet(x[train, ], y[train], alpha = 0, family = binomial)
plot(cv.out)

bestlam <- cv.out$lambda.min
bestlam

ridge.probs <- predict(ridge.mod, s = bestlam,newx = x[!train, ], type = "response")

ridge.coeffs <- predict(ridge.mod, s = bestlam,newx = x[!train, ], type = "coefficients")

```
The lambda chosen for the ridge regression is 0.0469.

```{r}
ridge.coeffs[ridge.coeffs == 0]
```
Ridge regression includes all predictors in the model and does not perform variable selection. Although it shrinks the regression coefficients toward zero, it does not set any coefficients exactly to zero. Thus, all predictors including dummy variables for categorical variables remain in the model.

## Brier score
```{r}
brier_test_ridge <- brier_score(as.numeric(hd.test) - 1, ridge.probs)
brier_test_ridge
```
The brier score of ridge regression model is 0.1128.

## ROC curve
```{r}
roc_obj_ridge <- roc(as.numeric(hd.test) - 1, ridge.probs[,1])
plot(roc_obj_ridge, print.auc=TRUE)
```

An AUC with value of 0.920 indicates better discrimination as compared to the glm.When randomly select an individual with heart disease and one without heart disease,the model assigns higher predicted probability of identifying heart disease to the affected individual 92% of the time.


## Calibration curve
```{r}
cali.curve.ridge<-valProbggplot(ridge.probs[,1], as.numeric(hd.test)-1)
cali.curve.ridge
```
The calibration intercept is 0.23, which deviates from the ideal value of 0, indicating that the model systematically underestimates the predicted probabilities. The calibration slope is 1.26, which is greater than 1, suggesting that the predicted probabilities are not extreme enough.

# Lasso Regression

```{r}
set.seed(100)
lasso.mod <- glmnet(x[train, ], y[train], alpha = 1, family = binomial)
cv.out_lasso <- cv.glmnet(x[train, ], y[train], alpha = 1, family = binomial)
plot(cv.out_lasso)

bestlam_lasso <- cv.out_lasso$lambda.min
bestlam_lasso 

lasso.probs <- predict(lasso.mod, s = bestlam_lasso,newx = x[!train, ], type = "response")
# check the selected features in lasso 
lasso.coefs <- predict(lasso.mod, s = bestlam_lasso, type = "coefficients")
```
The lambda chosen in lasso is 0.0199.

```{r}
lasso.coefs[lasso.coefs != 0]
lasso.coefs[lasso.coefs == 0]
```
LASSO regression performs variable selection by shrinking some coefficients exactly to zero. Variables with coefficients shrunk to zero are effectively excluded from the model, indicating that they contribute little to the predictive performance. The remaining variables with non-zero coefficients are considered to be informative and contribute to the model.
In this case, there are 8 predictors including dummy variables in the categorical predictor has been excluded

## Brier score
```{r}
brier_test_lasso <- brier_score(as.numeric(hd.test) - 1, lasso.probs)
brier_test_lasso
```
The brier score of lasso regression model is 0.1179.

## ROC curve
```{r}
roc_obj_lasso <- roc(as.numeric(hd.test) - 1, lasso.probs[,1])
plot(roc_obj_lasso, print.auc=TRUE)
```

An AUC with value of 0.912 indicates when randomly select an individual with heart disease and one without heart disease,the model assigns higher predicted probability of identifying heart disease to the affected individual 91.2% of the time.  

## Calibration curve
```{r}
cali.curve.lasso<-valProbggplot(lasso.probs[,1], as.numeric(hd.test)-1)
cali.curve.lasso
```
The calibration intercept is 0.23, which deviates from the ideal value of 0, indicating that the model systematically underestimates the predicted probabilities. The calibration slope is 1.15, which is greater than 1, suggesting that the predicted probabilities are not extreme enough.

# LDA - Linear Discriminant Analysis
```{r}
set.seed(100)
lda.fit <- lda(heart_disease ~ ., data = train.hd)
lda.fit

lda.pred <- predict(lda.fit,newdata=test.hd)
lda.pred.probs <- lda.pred$posterior[, 2]
```
Based on the prior probabilities of group 0 and 1, it has 0.5571 and 0.4429 respectively. It indicates that there are 44.29% individuals have heart disease in training dataset.

The coefficients of linear discriminant indicates that variable have positive coefficients, suggesting that higher values of these variable contribute positively to the discriminant score and associate with higher likelihood of heart disease. In this scenario, only variable age has the negative coefficients.


## Brier score
```{r}
brier_test_lda <- brier_score(as.numeric(hd.test) - 1, lda.pred.probs)
brier_test_lda
```
The brier score of lda model is 0.1152.

## ROC curve
```{r}
roc_obj_lda <- roc(as.numeric(hd.test) - 1, lda.pred.probs)
plot(roc_obj_lda, print.auc=TRUE)
```

An AUC with value of 0.909 indicates when randomly select an individual with heart disease and one without heart disease,the model assigns higher predicted probability of identifying heart disease to the affected individual 90.9% of the time.

## Calibration curve
```{r}
cali.curve.lda<-valProbggplot(lda.pred.probs, as.numeric(hd.test)-1)
cali.curve.lda
```
The calibration intercept is 0.49, which deviates from the ideal value of 0, indicating that the model systematically underestimates the predicted probabilities. The calibration slope is 0.65, which is less than 1, suggesting that the predicted probabilities are too extreme.

# Naive Bayes
```{r}
set.seed(100)
nb.fit <- naiveBayes(heart_disease ~ ., data = train.hd)
nb.fit
nb.preds <- predict(nb.fit, test.hd, type = "raw")
```
In the Naive Bayes model, it shows the mean and standard deviation of all covariates when heart disease is either 0 or 1.

## Brier Score
```{r}
brier_test_nb<- brier_score(as.numeric(hd.test) - 1, nb.preds[,2])
brier_test_nb
```
The brier score of naive bayes model is 0.1111.

## ROC curve
```{r}
roc_obj_nb <- roc(as.numeric(hd.test) - 1, nb.preds[,2])
plot(roc_obj_nb, print.auc=TRUE)
```
An AUC with value of 0.932 indicates when randomly select an individual with heart disease and one without heart disease,the model assigns higher predicted probability of identifying heart disease to the affected individual 93.2% of the time.

## Calibration curve
```{r}
cali.curve.nb <- valProbggplot(nb.preds[,2], as.numeric(hd.test)-1)
cali.curve.nb
```
The calibration intercept is 0.48, which deviates from the ideal value of 0, indicating that the model systematically underestimates the predicted probabilities. The calibration slope is 0.53, which is less than 1, suggesting that the predicted probabilities are too extreme.

# Trees
```{r}
set.seed(100)
tree.fit <- tree(heart_disease ~ ., data = train.hd)
summary(tree.fit)
```
It shows that the training misclassification error rate is 9.05%.


```{r}
plot(tree.fit)
text(tree.fit,pretty =0,,cex=0.6)
```
It shows that the most important indicator of 'heart disease' is cp, since the first branch differentiates between individuals with cp<3.5 and cp>=3.5

**Pruning the tree**
```{r}
set.seed(10)
cv_tree <- cv.tree(tree.fit,FUN = prune.misclass)
cv_tree
```
The tree model with 12 terminal nodes results in the minimum,42 cross validation errors.

```{r}
par(mfrow = c(1, 2))
plot(cv_tree$size, cv_tree$dev, type = "b")
plot(cv_tree$k, cv_tree$dev, type = "b")
```
```{r}
prune.tree <- prune.misclass(tree.fit, best = 12)
plot(prune.tree)
text(prune.tree, pretty = 0,cex=0.6)
```
```{r}
tree.preds <- predict(prune.tree,newdata = test.hd)[,2]
```



## Brier Score
```{r}
brier_test_tree<- brier_score(as.numeric(hd.test) - 1, tree.preds)
brier_test_tree
```
The brier score of naive bayes model is 0.1569.

## ROC curve
```{r}
roc_obj_tree <- roc(as.numeric(hd.test) - 1, tree.preds)
plot(roc_obj_tree, print.auc=TRUE)
```
An AUC with value of 0.857 indicates when randomly select an individual with heart disease and one without heart disease,the model assigns higher predicted probability of identifying heart disease to the affected individual 85.7% of the time.

## Calibration curve
```{r}
cali.curve.tree <- valProbggplot(tree.preds, as.numeric(hd.test)-1)
cali.curve.tree
```
The calibration intercept is 0.83, which deviates from the ideal value of 0, indicating that the model systematically underestimates the predicted probabilities. The calibration slope is 0.81, which is less than 1, suggesting that the predicted probabilities are too extreme.

# Random Forest
```{r}
set.seed(100)
bag.fit<- randomForest(heart_disease ~ ., data = train.hd, mtry = 13, importance = TRUE)
bag.fit

yhat.bag <- predict(bag.fit,newdata = test.hd,type="prob")[,2]
```

## Brier Score
```{r}
brier_test_bag<- brier_score(as.numeric(hd.test) - 1, yhat.bag)
brier_test_bag
```
The brier score of naive bayes model is 0.1348.

## ROC curve
```{r}
roc_obj_bag <- roc(as.numeric(hd.test) - 1, yhat.bag)
plot(roc_obj_bag, print.auc=TRUE)
```

An AUC with value of 0.888 indicates when randomly select an individual with heart disease and one without heart disease,the model assigns higher predicted probability of identifying heart disease to the affected individual 88.8% of the time.

## Calibration curve
```{r}
cali.curve.bag <- valProbggplot(yhat.bag, as.numeric(hd.test)-1)
cali.curve.bag
```
The calibration intercept is 0.24, which deviates from the ideal value of 0, indicating that the model systematically underestimates the predicted probabilities. The calibration slope is 0.92, which is less than 1, suggesting that the predicted probabilities are too extreme.




# Comparison on the roc curve for models
**roc**
```{r}
    roc_glm=data.frame(
      model="glm",
      spec=roc_obj_glm$specificities,
      sens=roc_obj_glm$sensitivities
      )
    roc_ridge=data.frame(
      model="ridge",
      spec=roc_obj_ridge$specificities,
      sens=roc_obj_ridge$sensitivities
      )
    roc_lasso=data.frame(
      model="lasso",
      spec=roc_obj_lasso$specificities,
      sens=roc_obj_lasso$sensitivities
    )
    roc_lda=data.frame(
      model="lda",
      spec=roc_obj_lda$specificities,
      sens=roc_obj_lda$sensitivities
      )
    roc_nb=data.frame(
      model="nb",
      spec=roc_obj_nb$specificities,
      sens=roc_obj_nb$sensitivities
    )
    roc_tree=data.frame(
      model="tree",
      spec=roc_obj_tree$specificities,
      sens=roc_obj_tree$sensitivities
    )
  roc_bag = data.frame(
      model="bagging",
      spec=roc_obj_bag$specificities,
      sens=roc_obj_bag$sensitivities
    )

roc_df <- rbind(roc_glm,roc_ridge,roc_lasso,roc_lda,roc_nb,roc_tree,roc_bag)
  
ggplot(data=roc_df,aes(x=1-spec,y=sens,color=model))+
  geom_line(size=0.8)+
  geom_abline(linetype = "dashed") +
  theme_minimal()+labs(
    x="1-Specificity",
    y="Sensitivity",
    title="ROC curve comparison"
  )
```

The plot shows an overall comparison of all models’ classification performance.


```{r}
ggplot() +
  geom_line(data = subset(roc_df, model != "nb"),
            aes(x = 1 - spec, y = sens, group = model),
            color = "grey", alpha = 0.5) +
  
  geom_line(data = subset(roc_df, model == "ridge"),
            aes(x = 1 - spec, y = sens, color = "Ridge Regression"),
            size = 0.8) +
  
  geom_line(data = subset(roc_df, model == "nb"),
            aes(x = 1 - spec, y = sens, color = "Naive Bayes"),
            size = 0.8) +
  
  scale_color_manual(
    name = "Model",
    values = c("Naive Bayes" = "red",
               "Ridge Regression" = "purple")
  ) +
  
  geom_abline(linetype = "dashed") +
  theme_minimal() +
  labs(
    x = "1 - Specificity",
    y = "Sensitivity",
    title = "ROC Curve Comparison on Top 2 Models"
  )
```
# Comparison of models performance on estimate of calibration curve,brier score and the AUC score
```{r}
 glm=data.frame(
      model="glm",
      slope=cali.curve.glm$Calibration$Slope[[1]],
      brier=brier_test_glm,
      auc=roc_obj_glm$auc
      )
  ridge=data.frame(
      model="ridge",
      slope=cali.curve.ridge$Calibration$Slope[[1]],
      brier=brier_test_ridge,
      auc=roc_obj_ridge$auc
      )
  lasso=data.frame(
      model="lasso",
      slope=cali.curve.lasso$Calibration$Slope[[1]],
      brier=brier_test_lasso,
      auc=roc_obj_lasso$auc
    )
  lda=data.frame(
      model="lda",
      slope=cali.curve.lda$Calibration$Slope[[1]],
      brier=brier_test_lda,
      auc=roc_obj_lda$auc
      )
  nb=data.frame(
      model="nb",
      slope=cali.curve.nb$Calibration$Slope[[1]],
      brier=brier_test_nb,
      auc=roc_obj_nb$auc
    )
  tree=data.frame(
      model="tree",
      slope=cali.curve.tree$Calibration$Slope[[1]],
      brier=brier_test_tree,
      auc=roc_obj_tree$auc
    )
  bag = data.frame(
      model="bagging",
      slope=cali.curve.bag$Calibration$Slope[[1]],
      brier=brier_test_bag,
      auc=roc_obj_bag$auc
    )

comparison_df <- rbind(glm,ridge,lasso,lda,nb,tree,bag)

comparison_df
```
It shows that naive bayes(NB) model achieves the lowest brier score of 0.1122 and the highest AUC score of 0.9329. This indicates better discrimination and low prediction error.However, its calibration slope of 0.5277 is far from 1, suggesting a poor calibration and predict probabilities too extreme. 

In fact, bagging has relative lower AUC score of 0.8883 and higher brier score of 0.1569, indicates weaker discrimination, but it yields the calibration slope of 0.9221, suggesting its predicted probabilities are more aligned with the observed outcomes.


As a conclusion, bagging model may be preferred, where the primary objective is accuracy probability estimation. Whereas, Naive Bayes is more appropriate when the primary objective is discrimination. 
